---
title: "Slouch"
author: "Bjørn Tore Kopperud"
date: "`r Sys.Date()`"
fontsize: 12pt
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# User guide

Slouch is an implementation of a particular phylogenetic comparative method. It can fit univariate among-species Ornstein-Uhlenbeck models of phenotypic trait evolution, where the trait evolves toward a primary optimum. Optima can be fitted either as discrete regimes as niches on the phylogenetic tree, and/or with continuous covariates. For more information about the theoretical background and how the method is derived, see the following articles

- Hansen (1997)^[Hansen, T. F. (1997). Stabilizing Selection and the Comparative Analysis of Adaptation. Evolution, 51(5), 1341. https://doi.org/10.2307/2411186]
- Butler & King (2004)^[Butler, M. A., & King, A. A. (2004). Phylogenetic comparative analysis: a modeling approach for adaptive evolution. American Naturalist, 164(6), 683–695. https://doi.org/10.1086/426002]
- Hansen *et al.* (2008)^[Hansen, T. F., Pienaar, J., & Orzack, S. H. (2008). A comparative method for studying adaptation to a randomly evolving environment. Evolution, 62(8), 1965–1977. https://doi.org/10.1111/j.1558-5646.2008.00412.x]
- Labra *et al.* (2009)^[Labra, A., Pienaar, J., & Hansen, T. F. (2009). Evolution of Thermal Physiology in Liolaemus Lizards: Adaptation, Phylogenetic Inertia, and Niche Tracking. The American Naturalist, 174(2), 204–220. https://doi.org/10.1086/600088]
- Hansen & Bartoszek (2012)^[Hansen, T. F., & Bartoszek, K. (2012). Interpreting the evolutionary regression: The interplay between observational and biological errors in phylogenetic comparative studies. Systematic Biology, 61(3), 413–425. https://doi.org/10.1093/sysbio/syr122]

## Prerequisites

In order to fit an Ornstein-Uhlenbeck model in slouch, you will need to have a phylogenetic tree of interest in the `phylo` format, from package `ape`. The phylogenetic tree must be rooted. Polytomies and non-branching edges are allowed. Both ultrametric and non-ultrametric trees can be used, but this will have implications for what parameters can reliably be estimated. For the purposes of illustrating syntax, let's create an arbitrary tree.

```{r, fig.width=7}
library(ape)

set.seed(8)
n <- 80
phy <- rtree(n)

par(mar = c(0,0,0,0))
plot(ladderize(phy))
```

Slouch can fit univariate Ornstein-Uhlenbeck models. Let's create a toy dataset. The species in the data frame need to be in a particular order to line up with the order of the tree, so I will intentionally scramble the dataframe first, to illustrate how to reorder it.

```{r, fig.width = 7}
mydata <- data.frame(species = phy$tip.label,
                     y = rnorm(n, sd = 4) + 1:n*1.7,
                     x = rnorm(n, sd = 4) + 1:n,
                     z = rnorm(n, sd = 4) + 1:n*(-0.6))

plot(mydata$x, mydata$y, ylab = "y", xlab = "x")

## Scramble the rows
mydata <- mydata[sample(1:n, n),]

## Check whether they are lined up correctly
mydata$species == phy$tip.label
```
Unsurprisingly, not a single one of the species is in its correct place. As `slouch` is (for now) intentionally programmed in a *dumb* way, we will have to reorder the species. Here is one way to do it.

```{r}
mydata <- mydata[match(phy$tip.label, mydata$species) ,]

## Check if they line up again
mydata$species == phy$tip.label
```

## Single optimum model

Now we are ready to fit the first models, for which we will use the function `slouch.fit`. The following is an intercept only model, or a single optimum model. In `slouch`, there are two main techniques for estimating the most likely $\alpha$ and $\sigma_y^2$ parameters. In this particular implementation the $\sigma_y^2$ is reparameterized as $\frac{\sigma_y^2}{2\alpha}$, which is the variance of the predictor variable when it is stationary around the optimum. The first way is to use a grid-search of likely parameters from user input, which may take some trial and error.

```{r, fig.show='hold', fig.width=3}
library(slouch)
model0 <- slouch.fit(phy = phy,
                     half_life_values = seq(0, 50, length.out = 10),
                     vy_values = seq(150, 800, length.out = 10),
                     species = mydata$species,
                     response = mydata$y)
plot(model0)

model0 <- slouch.fit(phy = phy,
                     half_life_values = seq(0, 100, length.out = 10),
                     vy_values = seq(350, 2500, length.out = 10),
                     species = mydata$species,
                     response = mydata$y)
plot(model0)
```

These plots of the likelihood surfaces are both based on the same data, but with different grid locations. The optima and model fit statistics that are reported in the output are conditional on a particular combination of these $\alpha$ and $\frac{\sigma_y^2}{2\alpha}$. When the grid-search is used, these parameters are the combination which gives the highest log-likelihood; the peak of the surface in the likelihood plot. If the grid-search does not contain the true maximum likelihood, the model outputs will reflect this.

Another, perhaps more convenient way of estimating parameters is to use the hillclimber function. On default it will start on a random combination of $\alpha$ and $\frac{\sigma_y^2}{2\alpha}$, but this may also be specified. While the hillclimber might seem both faster and more accurate at first glance, there are some drawbacks. If the likelihood search space has one or more local maxima, the hillclimber may converge at a sub-optimal location and give parameter estimates which are not truly maximum likelihood estimates.

```{r}
model0 <- slouch.fit(phy = phy,
                     species = mydata$species,
                     response = mydata$y,
                     hillclimb = TRUE)
plot(model0)
model0$oupar
```

In this case, the ML estimate of the phylogenetic half-life $t_{1/2} = \frac{log(2)}{\alpha}$ was very large. The units of the phylogenetic half-life are the same units as the branch lengths in the phylogenetic tree, `phy$edge.length`. The total depth, or distance from the root, can for all nodes be calculated with `node.depth.edgelength(phy)`. In this case the maximum depth is about 6.65. Given that the estimated half-life $t_{1/2}$ was many times larger than the total length of the phylogeny, and that $\alpha$ is very close to zero, we can say that there is next to no strength of pull toward the optimum. If there is no such pull, the model collapses to a Brownian motion.

## Optima as a linear regression

Slouch can also fit continuous covariates to the model, with the optimum being a linear regression. The covariates may either be fitted as direct effects without a phylogenetic covariance structure, or as univariate Brownian motion variables. Let's fit our toy x as a direct effect. One problem with fitting direct effect covariate models is that its residual variance-covariance matrix collapses if $\frac{\sigma_y^2}{2\alpha}$ reaches zero. The immediate consequence is that the residual variance-covariance matrix $V$ is singular, non-invertible, and the program will crash. If within-species observational error is added to the model, this does not happen. In order to use the hillclimber in this scenario, it may be necessary to constrain its search space as to not enter zero or close to zero. The exact feasible boundary for this may depend on the scale of the response trait.

```{r, fig.show='hold'}
model1 <- slouch.fit(phy = phy,
                     species = mydata$species,
                     response = mydata$y,
                     fixed.cov = mydata$x,
                     lower = c(0, 0.01),
                     hillclimb = TRUE)
plot(model1)

plot(mydata$x, mydata$y, ylab = "y", xlab = "x", main = "Trait plot")
abline(lm(mydata$y ~ mydata$x), col = "black", lwd = 2)
abline(model1$opt.reg$coefficients[,1], col = "orange", lwd = 2)
model1$opt.reg$coefficients
```

Even though the single-optimum model showed considerable phylogenetic inertia, in this model it has completely vanished. In this case, the regression line estimated using ordinary least squares is indistinguishable from the Ornstein-Uhlenbeck optimal regression. Given that we started with random normally distributed variables without any phylogenetic correlation structure, this makes sense. It is also possible to fit a model with multiple continuous covariates, however the input to `fixed.cov` must be a matrix that has column names.

```{r}
model2 <- slouch.fit(phy = phy,
                     species = mydata$species,
                     response = mydata$y,
                     fixed.cov = cbind(x = mydata$x, z = mydata$z),
                     lower = c(0, 0.01),
                     hillclimb = TRUE)
plot(model2)
```




